---
title: "Lab_9"
author: "Aileen Andrade-Torres"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: paged
        theme: cosmo
        toc: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "build") })
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Setup
library(pacman) 
library(DBI)
library(summarytools)
library(ggplot2)
p_load("tidyverse", "knitr")
```         

## Responses  
### Question 2: I believe all of these variables could be important for the analysis. If a patient refused to answer or didn’t know, that suggests the missingness is not random. But if a value is simply marked as missing, we don’t know the reason behind it. I’d want a clearer understanding of what’s missing and why before determining how to handle it.


```{r part_1, echo=FALSE, message=FALSE, warning=FALSE}                 
# code for part 1
## Question 1 - Reading om SQL data
db <- dbConnect(RSQLite::SQLite(), dbname = "epi560.sqlite")
audiometry <- dbGetQuery(db, "SELECT * from P_AUQ")
sleep <- dbGetQuery(db, "SELECT * from P_SLQ")

## Merging 
data <- dbGetQuery(db, "SELECT * FROM P_SLQ LEFT JOIN P_AUQ ON
                   P_SLQ.SEQN = P_AUQ.SEQN")
#View(data)

## Question 3 
data_clean <- data

### Frequency before cleaning
freq(data_clean$SLQ300, cum = TRUE, total = TRUE, na.last = TRUE)
freq(data_clean$SLQ310, cum = TRUE, total = TRUE, na.last = TRUE)

### Recode refused, don't know, and missing
#### SLQ300
data_clean$SLQ300[data_clean$SLQ300 == "77777"] <- NA
data_clean$SLQ300[data_clean$SLQ300 == "99999"] <- NA
data_clean$SLQ300[data_clean$SLQ300 == ""] <- NA

#### SLQ310
data_clean$SLQ310[data_clean$SLQ310 == "77777"] <- NA
data_clean$SLQ310[data_clean$SLQ310 == "99999"] <- NA
data_clean$SLQ310[data_clean$SLQ310 == ""] <- NA

###Frequency table after cleaning
freq(data_clean$SLQ300, cum = TRUE, total = TRUE, na.last = TRUE)
freq(data_clean$SLQ310, cum = TRUE, total = TRUE, na.last = TRUE)

## Question 4 - Cleaning more variables (SLQ300, SLQ310, SLQ320, SLQ330)
### Variables to be cleaned
var_clean <- c("SLQ300", "SLQ310", "SLQ320", "SLQ330")

### How to clean those variables
missing <- list(
  SLQ300 = c("77777", "99999", ""),
  SLQ310 = c("77777", "99999", ""),
  SLQ320 = c("77777", "99999", ""),
  SLQ330 = c("77777", "99999", "")
)

### Loop 
for (var in var_clean) {
  data_clean[[var]] <- as.character(data_clean[[var]])
  data_clean[[var]][data_clean[[var]] %in% missing[[var]]] <- NA
}

### Frequency table of SLQ320
freq(data_clean$SLQ320, cum = TRUE, total = TRUE, na.last = TRUE)

```


### Question 5: In the audiometry dataset, it’s important to assess how well the patient can hear, whether they have hearing loss, the severity of that loss, and possibly its cause. For sleep, knowing how much sleep the individual gets is key. Additional variables like medical conditions and demographic factors such as age may also be important, as they can influence both hearing loss and sleep patterns. 

#*Question 6*
```{r question 6, echo=FALSE, message=FALSE, warning=FALSE}

#join tables
data2 <- dbGetQuery(db, "
  SELECT *
  FROM P_SLQ
  LEFT JOIN P_PAQ ON P_SLQ.SEQN = P_PAQ.SEQN
  LEFT JOIN P_ALQ ON P_SLQ.SEQN = P_ALQ.SEQN
  LEFT JOIN P_GHB ON P_SLQ.SEQN = P_GHB.SEQN
  LEFT JOIN P_AUQ ON P_SLQ.SEQN = P_AUQ.SEQN
")

#create new data frame
data_clean2 <- data2

#cleaning specific variables

##Sleep disorders- SLQ050 missingness
data_clean2$SLQ050[data_clean2$SLQ050 == 7] <- NA
data_clean2$SLQ050[data_clean2$SLQ050 == 9] <- NA
data_clean2$SLQ050[data_clean2$SLQ050 == "."] <- NA

##Physical activity- PAQ610 missingness
data_clean2$PAQ610[data_clean2$PAQ610 == 77] <- NA
data_clean2$PAQ610[data_clean2$PAQ610 == 99] <- NA
data_clean2$PAQ610[data_clean2$PAQ610 == "."] <- NA

##Alcohol use- ALQ121 missingness
data_clean2$ALQ121[data_clean2$ALQ121 == 77] <- NA
data_clean2$ALQ121[data_clean2$ALQ121 == 99] <- NA
data_clean2$ALQ121[data_clean2$ALQ121 == "."] <- NA


##Audiometry-AUQ054 missingness
data_clean2$AUQ054[data_clean2$AUQ054 == 77] <- NA
data_clean2$AUQ054[data_clean2$AUQ054 == 99] <- NA
data_clean2$AUQ054[data_clean2$AUQ054 == "."] <- NA

##Glycohemoglobin-LBXGH missingness
data_clean2$LBXGH[data_clean2$LBXGH == "."] <- NA

kable(head(data_clean2[, c("SLQ050", "PAQ610", "ALQ121", "AUQ054", "LBXGH")]))
```


##*Question 7*
```{r question 7, echo=FALSE}

#join tables
data3 <- dbGetQuery(db, "
  SELECT *
  FROM P_SLQ
  LEFT JOIN P_PAQ ON P_SLQ.SEQN = P_PAQ.SEQN
  LEFT JOIN P_ALQ ON P_SLQ.SEQN = P_ALQ.SEQN
  LEFT JOIN P_GHB ON P_SLQ.SEQN = P_GHB.SEQN
  LEFT JOIN P_AUQ ON P_SLQ.SEQN = P_AUQ.SEQN
")

#Create functions

##continuous
clean_continuous <- function(data, var_name) {
  lower_percentile <- quantile(data[[var_name]], 0.05, na.rm = TRUE)
  upper_percentile <- quantile(data[[var_name]], 0.95, na.rm = TRUE)
  
  data[[var_name]][data[[var_name]] < lower_percentile] <- lower_percentile
  
  data[[var_name]][data[[var_name]] > upper_percentile] <- upper_percentile
  
  return(data)
}


#categorical
clean_categorical <- function(data, var_name) {
  data[[var_name]][data[[var_name]] == 99] <- 9999
  return(data)
}

#create new data frame
data_clean3 <- data3

#create vectors
cont_var <- c("LBXGH")
cat_var <- c("PAQ610", "ALQ121", "AUQ054", "SLQ050")

#apply functions-continuous

for (var in cont_var) {
  data_clean3 <- clean_continuous(data_clean3, var)
}

#apply functions-categorical
for (var in cat_var) {
  data_clean3 <- clean_categorical(data_clean3, var)
}

#display with kable
kable(head(data_clean3[, c("PAQ610", "ALQ121", "LBXGH", "AUQ054", "SLQ050")]))

```
#*Histograms* 

```{r histograms, echo=FALSE}
#original data
hist(data3$LBXGH, 
     main = "Histogram of Glycohemoglobin (%) - Original Data", 
     xlab = "Glycohemoglobin (%)", 
     col = "steelblue", 
     border = "orange", 
     breaks = 20) 

#cleaned data
hist(data_clean3$LBXGH, 
     main = "Histogram of Glycohemoglobin (%) - Cleaned Data", 
     xlab = "Glycohemoglobin (%)", 
     col = "aquamarine", 
     border = "yellow", 
     breaks = 20)

#end session

dbDisconnect(db)
```
 ### Chat GPT was used to help trouble shoot why Summarytools was not running in R. 

\pagebreak  

## Code Appendix  

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE} 
```
